---
title: "Get started with pika"
author: "Kylie Ainslie"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with pika}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`pika` answers two questions about a pair of time series, optionally repeated across
a grouping variable (e.g. region): at what lag are they most strongly correlated,
and how does that correlation change over time? This page is a fast tour of the
core functions on the data bundled with the package. For two full worked examples
with real data and discussion, see the case studies:
`vignette("china-mobility-rt")` and `vignette("wastewater-surveillance")`.

## Installation

```r
# install.packages("remotes")
remotes::install_github("mrc-ide/pika")
```

## Lag and rolling correlation
`pika` bundles weekly SARS-CoV-2 wastewater concentration and COVID-19 case counts
for three US states (`wastewater_data` and `covid_case_data`). Joining them gives a
pair of time series per state.

```{r setup_data, message=FALSE}
library(dplyr)
library(pika)

data(wastewater_data)
data(covid_case_data)

dat <- inner_join(wastewater_data, covid_case_data, by = c("state", "date")) %>%
  mutate(log_conc = log(conc), log_cases = log(cases + 1))
```

`cross_corr()` finds the lag (in rows of your data -- here, weeks) at which two
variables are most strongly correlated, separately for each level of `grp_var`:

```{r cross_corr}
lags <- cross_corr(
  dat = dat, date_var = "date", grp_var = "state",
  x_var = "log_conc", y_var = "log_cases", max_lag = 8
)
lags
```

`rolling_corr()` then calculates the correlation between the two variables in a
moving window, so you can see whether that relationship is stable over time rather
than assuming a single lag/correlation applies throughout:

```{r rolling_corr}
dat_corr <- rolling_corr(
  dat = dat, date_var = "date", grp_var = "state",
  x_var = "log_conc", y_var = "log_cases", n = 12
)
tail(dat_corr, 3)
```

`plot_corr()` visualizes both series alongside the rolling correlation, faceted by
group:

```{r plot_corr, fig.width=8, fig.height=6, message=FALSE, warning=FALSE}
plot_corr(
  dat = dat_corr, date_var = "date", grp_var = "state",
  x_var = "log_conc", y_var = "log_cases"
)
```

`plot_lag()` plots a histogram of the lags returned by `cross_corr()` across
groups, which is more useful with many groups than the three shown here:

```{r plot_lag, fig.width=5, fig.height=3}
plot_lag(dat = lags, lag_var = "lag")
```

## Other functions

`calc_percent_change()` converts a count variable into percent change relative to
the average value in a baseline period:

```{r percent_change}
pct <- calc_percent_change(
  dat = wastewater_data, date_var = "date", grp_var = "state",
  count_var = "conc", n_baseline_periods = 4
)
tail(pct, 3)
```

`estimate_rt()` wraps `EpiEstim::estimate_R()` to estimate the effective
reproduction number (Rt) from case or death counts, by group:

```{r estimate_rt, message=FALSE, warning=FALSE}
data(china_case_data)
rt <- estimate_rt(
  dat = china_case_data, grp_var = "province",
  date_var = "date", incidence_var = "cases"
)
tail(rt, 3)
```

## Learn more
For a full worked example that estimates Rt and determines its lagged relationship
with mobility, reproducing published results, see `vignette("china-mobility-rt")`.
For an example that tests whether wastewater surveillance is a reliable leading
indicator for clinical case counts, see `vignette("wastewater-surveillance")`.
