---
title: "pika: A lightweight package for evaluating correlation between time series"
author: "Kylie Ainslie"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: pika_vignette.bib
csl: plos.csl
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview
`pika` is a lightweight package that enables easily estimating the optimal lag and rolling ocrrleation between two time series. This package was dveloped for use with epidemiological data and mobility data, but can be used more generally for other time series. The main utility of this package is the ability to determine quantities of interest by a grouping variable, such as the rolling correlation between two time series for different locations. This vignette goes over the different functionality of `pika` in an epidemiological context, namely determining the optimal lag and rolling correlation between estimates of reproduction number (the average number of new cases generated by a single case) and mobility within the population within different regions of China during the COVID-19 pandemic.

### Data
`pika` contains several example data sets in the `data/` directory that can be used to run through the package functions. The example data sets include
1. Daily confirmed cases of COVID-19 from different provinces in China (china_case_data)
2. Estimates of effective reproduction number (Rt) over time using the case data in 1 (china_rt_estimates)
3. Daily within-city movement data for the same  provinces in 1 (exante_movement_data)

The case data has daily confirmed confirmed cases for different provinces in China from 16 January to 24 March 2020 from the dashboard maintained by Chinese Center for Disease Prevention and Control (CCDC) [@noauthor_2020-pv]. The CCDC dashboard collates numbers of confirmed cases reported by national and local health commissions in each province in mainland China, and Hong Kong SAR and Macau SAR. Confirmed cases are defined as suspected cases, who have epidemiological links and/or clinical symptoms, and are detected with SARS-CoV-2 by PCR tests. However, in Hubei province, clinically diagnosed cases were additionally included between 12 and 19 February.

`pika` provides a function (`estimate_rt()`) that enables users to input case data similar to that in 1 and estimate Rt by group using `EpiEstim`. However, for generalisability, a data set of Rt estimates is provided as the primary time series on which to perform the correlation analysis. The Rt estimates were estimated using `pika`'s `estimate_rt()` with `si_mean = 6.48` and `si_std = 3.83` [@Ferguson_undated-yc].

The daily within-city movement data, used as a proxy for economic activity, is available from 1 January to 24 March 2020 for major metropolitan cities within each province in mainland China, Hong Kong SAR, and Macau SAR. These data, provided by Exante Data Inc [@Exante_Data_Inc2020-mo], measured travel activity relative to the 2019 average (excluding Lunar New Year). The underlying data are based on near real-time people movement statistics from Baidu.

```{r load_data, echo=FALSE, message=FALSE}
# Load required packages ----------------------------------------------------------------
library(dplyr)
library(tidyr)
library(TTR)
library(EpiEstim)
library(ggplot2)
library(scales)
library(RColorBrewer)

# Load pika -----------------------------------------------------------------------------
devtools::load_all("~/Documents/pika")

# Load data -----------------------------------------------------------------------------
# load("data/china_rt_estimates.rda")
# load("data/exante_movement_data.rda")
# load("data/china_case_data.rda")

```

### Estimate effective reproduction number
To estimate the effective reproduction number (Rt), pika has a function `estimate_rt()` that provides a wrapper for `EpiEstim::estimate_R()` [@Cori2019-sc] that enables users to estimate Rt by a grouping variable. As input `estimate_rt()` takes a data frame of daily cases or deaths by group. `pika` includes an example data set `china_case_data.rda` that includes daily reported cases of COVID-19 from several provinces in China between 16 January 2020 and 24 March 2020.
```{r show_data, echo=FALSE, results='asis'}
knitr::kable(head(china_case_data, 10))
```

To estimate Rt from case data, the user must specify the data set in the `dat = ` argument of `estimate_rt()`. The data set should include three columns: a date column (this should be of class 'date'), a grouping variable, and the value used to estimate Rt (most commonly this is cases, but other quantities can be used, such as deaths). The names of the columns in the input data set should be specified as character strings in the `grp_var = `, `date_var = `, and `incidence_var = ` arguments.

```{r estimate_rt, message=FALSE, warning=FALSE}
# to estimate them with pika, use estimate_rt() ------------------------------------------
pika_rt_estimates <- estimate_rt(dat = china_case_data,
                                 grp_var = "province",
                                 date_var = "date",
                                 incidence_var = "cases"
                                 )
```

```{r rt_output, echo=FALSE}
knitr::kable(head(pika_rt_estimates, 10))
```

The method used to estimate Rt can be changed using the `est_method = ` argument. This argument takes the following options: `c("non_parametric_si","parametric_si","uncertain_si","si_from_data","si_from_sample")`. For methods that require specifying the mean and standard deviation of the serial interval, these parameters can be set using the `si_mean = ` and `si_std = ` function arguments.

For more information on the methodology underlying `EpiEstim::estimate_R` look at this [vignette](https://cran.r-project.org/web/packages/EpiEstim/vignettes/demo.html) or at `help("estimate_R")`.

## Lag
When determining the relationship between two time series it is sometimes of interest to determine the lag at which the correlation between the two time series is highest, we'll call this the optimal lag. To determine the optimal lag, the two time series must first be merged into a single data frame. We do this below using `dplyr`'s 'left_join()` by `date` and `province`.
```{r djoin_data, warning=FALSE, results='hide'}
# Join data sets together by date and province to determine cross correlation -----------
# for this we are using input rt estimates, not those estimated using pika::estimate_rt()
data_joined <- left_join(china_rt_estimates,
                         exante_movement_data,
                         by = c("date","province")
                         )
```

```{r, determine_lag}
# Determine lag with max cross correlation between Rt and movement ----------------------
lags <- cross_corr(dat = data_joined,
                   date_var = "date",
                   grp_var = "province",
                   x_var = "r_mean",
                   y_var = "movement",
                   max_lag = 10,
                   subset_date = '2020-02-15'
                  )

# use min lag across groups -------------------------------------------------------------
my_lag <- min(lags$lag)

```

```{r lag_var}
# create lag date using max lag from cross_corr() ---------------------------------------
data_joined_lag <- china_rt_estimates %>%
  mutate(date = date + my_lag) %>%
  left_join(., exante_movement_data, by = c("date", "province"))
```

## Rolling Correlation
```{r corr}
# Determine rolling correlation between Rt and movement ---------------------------------
data_corr <- rolling_corr(dat = data_joined_lag,
                          grp_var = "province",
                          x_var = "r_mean",
                          y_var = "movement",
                          n = 14)

```

## Figures
Plot of the 

```{r, fig.height=6, fig.width=8,fig.show='hold', message=FALSE, warning=FALSE}
# Plot Rt, movement, and correlation ----------------------------------------------------
my_labels <- c("beijing" = "Beijing", "guandong" = "Guangdong", "henan" = "Henan",
               "hong_kong_sar" = "Hong Kong SAR", "hubei" = "Hubei", "hunan" = "Hunan",
               "zhejiang" = "Zhejiang")
my_legend = c("Correlation", "Reproduction Number", "Movement")

plot_corr(dat = data_corr,
          date_var = "date",
          grp_var = "province",
          x_var = "r_mean",
          y_var = "movement",
          x_var_lower = "r_q2.5",
          x_var_upper = "r_q97.5",
          facet_labels = my_labels,
          legend_labels = my_legend,
          y_max = 10
          )
# Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.
```


## References

