---
title: "Case study: wastewater surveillance as a leading indicator for COVID-19 cases"
author: "Kylie Ainslie"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: wastewater-surveillance.bib
csl: plos.csl
vignette: >
  %\VignetteIndexEntry{Case study: wastewater surveillance as a leading indicator for COVID-19 cases}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview
Since SARS-CoV-2 RNA sheds into wastewater from both symptomatic and asymptomatic
infections, wastewater concentration has been proposed as a leading indicator for
clinical case counts that is unaffected by testing behaviour [@Peccia2020]. This
motivates a specific question that `cross_corr()` and `rolling_corr()` are built to
answer: **by how many days or weeks does the wastewater signal lead reported cases,
and did that lead time and the strength of the relationship hold up once clinical
case counts became a less reliable measure of true infections** (most notably after
at-home rapid antigen tests became widely available in the US in early 2022, after
which a growing share of infections were never reported to public health
authorities)?

This case study demonstrates the same lag-detection and rolling-correlation
workflow as `vignette("china-mobility-rt")`, applied to real, public health
surveillance data instead of the reproduction-number/mobility data used there. It is
a genuine test of the hypothesis above, not a constructed example, so the results
below are whatever the data show -- including where they complicate a tidy
narrative.

### Data
`pika` bundles two example datasets for this case study, covering California, New
York, and Ohio:

1. `wastewater_data`: weekly, state-level SARS-CoV-2 concentration in wastewater,
   built from site-level sample records published by the CDC National Wastewater
   Surveillance System (NWSS) [@NWSS]. Each state-week value is a
   population-weighted mean of the flow- and population-normalized concentration
   (copies per person per day) across all sites reporting that week.
2. `covid_case_data`: weekly newly reported COVID-19 cases for the same three
   states, built from cumulative case counts published by The New York Times
   [@NYTimes2023].

Both series run from early 2020 to 27 March 2023, the date the NYT repository
stopped updating -- which is also why this case study cannot look further forward
than that.

```{r load_data, message=FALSE}
library(dplyr)
library(knitr)
library(pika)

data(wastewater_data)
data(covid_case_data)
```

```{r show_data, echo=FALSE, results='asis'}
kable(head(wastewater_data, 5), caption = "wastewater_data")
kable(head(covid_case_data, 5), caption = "covid_case_data")
```

The number of sites contributing to each state-week grew substantially as NWSS
expanded, from a single site in California in April 2020 to well over 100 sites per
state by 2022-2023:

```{r n_sites_plot, fig.width=7, fig.height=3}
library(ggplot2)
ggplot(wastewater_data, aes(x = date, y = n_sites, colour = state)) +
  geom_line() +
  labs(x = NULL, y = "number of sites reporting", colour = NULL) +
  theme_minimal()
```

Concentration spans several orders of magnitude, so we work on the log scale
throughout, and similarly log-transform cases (adding 1 first, since some
state-weeks have zero newly reported cases early in the pandemic):

```{r join_data}
data_joined <- inner_join(wastewater_data, covid_case_data, by = c("state", "date")) %>%
  mutate(log_conc = log(conc), log_cases = log(cases + 1))
```

## Determining the optimal lag
`cross_corr()` searches over lags (in units of *rows* of the input data -- here,
weeks, since both series are already aggregated to one row per state per week) and
returns, for each group, the lag at which the cross-correlation between `x_var` and
`y_var` is highest, restricted to lags at or before zero (i.e. `x_var` at or ahead
of `y_var`). We treat wastewater concentration as `x_var` and cases as `y_var`, so a
negative lag means wastewater leads cases.

```{r cross_corr}
lags <- cross_corr(
  dat     = data_joined,
  date_var = "date",
  grp_var  = "state",
  x_var    = "log_conc",
  y_var    = "log_cases",
  max_lag  = 8
)
lags
```

```{r kable_lags, echo=FALSE}
kable(lags)
```

Wastewater leads cases by one to two weeks in California and Ohio, and is roughly
contemporaneous with cases in New York -- consistent with the days-to-weeks lead
times reported elsewhere in the wastewater surveillance literature. To align the two
series on this optimal lag, we shift each state's wastewater dates forward by
`-lag` weeks (i.e. `lag` rows, converted to days since our data is weekly).
**The lag returned by `cross_corr()` is in units of rows of the input data, not
necessarily days or weeks -- always check the spacing of your own data before
converting.**

```{r lag_var}
data_lagged <- data_joined %>%
  left_join(lags, by = "state") %>%
  mutate(date = date - lag * 7) %>%
  select(state, date, log_conc) %>%
  inner_join(
    data_joined %>% select(state, date, log_cases, cases),
    by = c("state", "date")
  ) %>%
  arrange(state, date)
```

## Rolling correlation
With the series aligned on their optimal lag, `rolling_corr()` calculates the
Pearson correlation between wastewater concentration and cases in a moving window,
which lets us see whether the strength of the relationship is stable over time
rather than assuming a single lag/correlation estimated across the whole series
applies throughout. We use an `n = 12` (twelve-week, roughly one quarter) window, a
compromise between smoothing out single-week noise and still resolving change over
the three years of data.

```{r corr}
data_corr <- rolling_corr(
  dat      = data_lagged,
  date_var = "date",
  grp_var  = "state",
  x_var    = "log_conc",
  y_var    = "log_cases",
  n        = 12
)
```

```{r plot_corr, fig.height=7, fig.width=8, fig.show='hold', message=FALSE, warning=FALSE}
plot_corr(
  dat          = data_corr,
  date_var     = "date",
  grp_var      = "state",
  x_var        = "log_conc",
  y_var        = "log_cases",
  legend_labels = c("Correlation", "log(wastewater conc.)", "log(cases + 1)")
)
```

## Did the relationship hold up as case-reporting quality declined?
At-home rapid testing became widely available in the US from around January 2022,
after which an increasing share of infections were never reported to public health
authorities. If that eroded the reliability of case counts as a measure of true
infection burden, we would expect the rolling correlation between wastewater and
cases to trend downward after early 2022.

```{r pre_post_summary}
data_corr %>%
  filter(!is.na(roll_corr)) %>%
  mutate(period = ifelse(date < as.Date("2022-03-01"), "pre-2022-03", "post-2022-03")) %>%
  group_by(state, period) %>%
  summarise(
    mean_corr   = round(mean(roll_corr), 2),
    median_corr = round(median(roll_corr), 2),
    .groups = "drop"
  ) %>%
  arrange(state, period) %>%
  kable()
```

Contrary to that expectation, the rolling correlation does **not** show a
systematic decline after March 2022 in any of the three states -- if anything, the
median correlation is similar or slightly higher in the later period. The rolling
correlation is visibly noisy throughout the whole series in the plot above, with
occasional weeks of weak or negative correlation in both periods, most likely
reflecting genuine week-to-week volatility in a 12-week window (holiday reporting
effects, changes in which sites reported that week) rather than a sustained
regime change.

This is a useful result precisely because it complicates the tidy assumption: at
the *state* level, aggregating across many counties and sewersheds, reported case
counts appear to have remained a reasonably proportional -- if biased -- signal of
underlying trends throughout the study period, even as absolute ascertainment
almost certainly fell. Whether the same holds at a finer geographic resolution
(individual sewersheds vs. county case counts) is a natural follow-up question, and
one where a single overall lag/correlation estimate -- rather than the rolling
version used here -- would have hidden the volatility entirely, and might have
supported the "relationship degraded" narrative or its opposite depending on
which weeks happened to dominate the estimate.

## References
